An Efficient Minibatch Acceptance Test for Metropolis-Hastings

### 2016-09-30 nips_2016.tex
Revised 6.2, Deleted Appendix C. Yeah, took too long. UP NEXT: Appendix A (UH OH...)

We conduct two sets of experiments to explore the benefits of our minibatch MH
test and to benchmark it with previous work. In Section~\ref{ssec:gaussians}, we
show that our test enables samples to converge to the posterior distribution of
a heated Gaussian mixture model. In Section~\ref{ssec:logistic}, we analyze its
efficiency on logistic regression.  Appendices~\ref{app:gaussian}
and~\ref{app:logistic} contain more detailed information on these respective
experiments.

@@

We next use logistic regression for the binary classification of 1s versus 7s in
the MNIST dataset~\cite{lecun-mnisthandwrittendigit-2010}. The data has 12007
training and 1000 testing elements (we used a random subset of the test data).
For the proposer, we use a random walk with covariance matrix $0.05I$ for the
$784\times 784$ identity matrix $I$. We set the posterior temperature at
$T=1000$ and use the minibatch size $m=100$ and compare
with~\cite{cutting_mh_2014} with tolerance 0.005 and
with~\cite{icml2014c1_bardenet14} with error bound control parameters $p = 2$
and $\delta = 0.01$.
% {\color{blue} Daniel: just observed this, but why are we
% using a subset of the test data? We really should use all 2000 (I think it's
% 2000, right?).}
% JFC- Would be good to fix, but not critical.

Figure~\ref{fig:logistic_performance} shows the prediction accuracy and log
likelihood on the test set as a function of the cumulative training data points
processed.  We see that our minibatch MH test is more efficient compared to the
other two methods, because it has similar or better performance while consuming
fewer data points.

@@

Figure~\ref{fig:logistic_minibatch} shows the histogram of minibatch sizes for
all three methods. With an initial minibatch size of 100, Korattikara 2014 and
Bardenet 2014 achieve average minibatch size of 2026 and 7612 respectively over
the MNIST classification task, while our method achieves that of 163, showing
significantly better performance than the benchmark test methods.

@@

\begin{itemize}[noitemsep]
    \item The proof of Lemma~\ref{lem:worst_case}.
    \item More information on Section~\ref{ssec:gaussians}.
    \item More information on Section~\ref{ssec:logistic}.
\end{itemize}

@@

In this section, we provide more details on Section~\ref{ssec:gaussians}.  Our
parameter is a 2-D vector
$\theta = \langle \theta_1, \theta_2 \rangle$, where
\begin{equation}
\theta_1 \sim \mathcal{N}(0, \sigma_1^2) \quad \mbox{ and } \quad \theta_2 \sim \mathcal{N}(0, \sigma_2^2)
\end{equation}
and where $\mathcal{N}$ indicates the normal distribution (more generally, the
multivariate normal). We consider the above as our prior.
Following~\cite{langevin_2011}, we set $\sigma_1^2 = 10$ and $\sigma_2^2=1$, so
the covariance matrix of $\theta$ is $\Sigma = {\rm diag}(10,1)$. Therefore, the
log prior probability we endow on $\theta$ is
\begin{equation}
\log p(\theta) = \log \left(\frac{1}{2\pi\sqrt{10}}\right) - \frac{1}{2}\theta^T\Sigma^{-1}\theta.
\end{equation}
To generate the data, we use the following Gaussian mixture with tied means:
\begin{equation}\label{eq:x_points}
x_i \sim \frac{1}{2}\mathcal{N}(\theta_1, \sigma_x^2) + \frac{1}{2}\mathcal{N}(\theta_1+\theta_2, \sigma_x^2)
\end{equation}
where, again following~\cite{langevin_2011}, we set $\sigma_x^2 = 2$. This means
the log likelihood of a single data instance is
\begin{equation}
\log p(x_i \mid \theta) = \log \frac{1}{4\sqrt{\pi}} +
\log\left(\exp\left(-\frac{1}{4}(x_i - \theta_1)^2\right) + \exp\left(-\frac{1}{4}(x_i -
(\theta_1+\theta_2))^2\right)\right)
\end{equation}
The official problem statement is as follows: given some number of conditionally
independent data points $x_1, x_2, \ldots, x_N$ generated according
to~(\ref{eq:x_points}), determine the posterior distribution of $\theta$:
\begin{equation}\label{eq:log_post}
\log p(\theta \mid x_1,\ldots,x_N) = \log p(\theta) + \sum_{i=1}^N\log p(x_i \mid \theta).
\end{equation}
Alternatively, depending on the application, we may opt to instead pick a point
estimate\footnote{We do this for the logistic regression experiment, described
in Section~\ref{ssec:logistic} and elaborated in Appendix~\ref{app:logistic}.}
of $\theta$, such as the MAP estimate. (If $N$ is extremely large, it will cause
the posterior to peak sharply at its modes, reducing distribution estimates to
point estimates.)

@@

\section{Logistic Regression Experiment Details}\label{app:logistic}

In this section, we go over some details of our logistic regression experiment. The feature vector
for an image consists of its pixel values, normalized between 0 and 1. For simplicity, we only
consider the binary classification case, so we only use digits one (denoted as output $y = 1$) and
seven (denoted as $y = -1$). The probability of the $i^{\rm th}$ output $y_i \in \{-1,+1\}$ with the
input vector $x_i$ is
\begin{equation}
p(y_i \mid x_i) = \frac{1}{1+\exp (-y_i \theta^T x_i)},
\end{equation}
where $\theta$ is the 784-length parameter vector.

For our experiment, we impose a uniform prior to represent our lack of knowledge about $\theta$.  We
use a random walk proposer, which can be modeled as $\theta' = \theta_i + \mathcal{N}(0,
\sigma^2I)$, where $\theta_i$ is the current sample, $\theta'$ is the proposed sample, and we choose
the variance to be a constant $\sigma^2 = 0.05$ for all components. We initialize $\theta_0$ to be a
vector of all ones, and set our minibatch size as $m=100$.

For our minibatch MH test, in order to enforce the ${\rm std}(\Delta') < 1.2$ condition, we use a
constant temperature $T = 1000$. If our estimated ${\rm std}(\Delta') \ge 1.2$, we ignore the
current iteration.  %Figure~\ref{fig:appendixExp2} plots our estimated ${\rm std}(\Delta')$ values
%versus iteration count.

%\begin{figure}[t]
  %\centering
  %\includegraphics[width=1\linewidth]{./figures/appendixExp2}
 % \caption{Additional results for the logistic regression experiment.}
  %\label{fig:appendixExp2}
%\end{figure}

%For adaptive MH testing, our experimental settings are the same as with our MH test, except we do
%not impose a temperature. The minibatch size of adaptive MH testing is also initialized as 50, but
%it may increase by that amount each iteration. Figure~\ref{fig:appendixExp2} shows the histogram of
%the actual minibatch size at the end of each iteration in adaptive MH testing.

\iffalse
\section{Neural Network Experiment Details}\label{app:nnet}

For our neural network experiment, we use the architecture discussed in Section~\ref{ssec:nets}. We
use the SGHMC as the proposer for our minibatch MH testing. For the baseline, we use a tuned
adaptive gradient descent optimizer, whose step size changes by $0.01 /(i+1)^{0.4}$. Both methods
have minibatch sizes set at $m=200$. There are one million total data instances $x_i$.

For SGHMC, we use the simplified update equations~\cite{sghmc_2014}:
\begin{equation}
\Delta \theta = v, \quad \quad\Delta v = -\eta \nabla U(\theta) - \alpha v + \mathcal{N}(0, 2(\alpha
-\hat{\beta}) \eta),
\end{equation}
where $v$ represents auxiliary momentum variables, and $\nabla U(\theta)$ is the gradient of the
system. We set the hyperparameters to be $\eta_i = 0.01 /(i+1)^{0.4}$, and $\alpha=0.1$. We use the
empirical Fisher~\cite{conf/icml/AhnBW12} information $V(\theta)$ to estimate the value of
$\hat{\beta}$, so that $\hat{\beta}=\frac{1}{2}\eta V(\theta)$. In order to control ${\rm
std}(\Delta')$, we initialize the temperature at 1000, and adjust it at iteration $i$ according to
$T_i = \max\{1, 1000/(i+1)^{0.5}\}$.

\fi

>>>>

We conduct two sets of experiments to explore the benefits of our minibatch MH
test and to benchmark it with previous work. In Section~\ref{ssec:gaussians}, we
show that our test enables samples to converge to the posterior distribution of
a heated Gaussian mixture model (additional details are in
Appendix~\ref{app:gaussian}). In Section~\ref{ssec:logistic}, we analyze its
efficiency on logistic regression.

@@

@@

We next use logistic regression for the binary classification of 1s versus 7s in
the MNIST dataset~\cite{lecun-mnisthandwrittendigit-2010}. The data has 12007
training and 1000 testing elements (we used a random subset of the test data).
The pixel values are scaled to be in $(0,1)$.  We impose a flat prior on
$\theta$. For the proposer, we again use a random walk, this time with
covariance matrix $0.05I$ for the $784\times 784$ identity matrix $I$. The
posterior temperature is set at $T=1000$.  We run our MH test and compare with
Korattikara 2014 and Bardenet 2014; minibatch sizes and other parameters are set
to be the same value as with the Gaussian Mixtures experiment from
Section~\ref{ssec:gaussians}.

Figure~\ref{fig:logistic_performance} shows the training log likelihood and the
testing prediction accuracy, each as a function of the cumulative training data
points processed.\footnote{Note that the curves do not span the same length over
the x-axis, because our test consumes fewer samples throughout the MCMC
procedure, so the corresponding curve will ``end'' before the other two.} To
generate the curves, for each of the 5000 sampled vectors $\theta_t$,
$t\in\{1,\ldots,5000\}$, we use $\theta_t$ as the parameter for logistic
regression.  We see that our minibatch MH test is more efficient compared to the
other two methods, because it achieves similar or superior performance while
consuming fewer data points.

Figure~\ref{fig:logistic_minibatch}, in a similar manner as
Figure~\ref{fig:gauss_mix_2}, shows the histogram of minibatch sizes for all
three methods on a log-log scale. With an initial minibatch size of 100,
Korattikara 2014 and Bardenet 2014 achieve average minibatch size of 2026 and
7612 respectively over the MNIST classification task, while our method achieves
that of 163, showing significantly better performance than the benchmark test
methods.

@@

\begin{itemize}[noitemsep]
    \item The proof of Lemma~\ref{lem:worst_case}.
    \item More information on Section~\ref{ssec:gaussians}.
\end{itemize}

@@

In this section, we provide more details on Section~\ref{ssec:gaussians}.  Our
parameter is a 2-D vector
$\theta = \langle \theta_1, \theta_2 \rangle$, where
\begin{equation}
\theta_1 \sim \mathcal{N}(0, \sigma_1^2) \quad \mbox{ and } \quad \theta_2 \sim \mathcal{N}(0, \sigma_2^2)
\end{equation}
and where $\mathcal{N}$ indicates the normal distribution (more generally, the
multivariate normal). We consider the above as our prior.
Following~\cite{langevin_2011}, we set $\sigma_1^2 = 10$ and $\sigma_2^2=1$, so
the covariance matrix of $\theta$ is $\Sigma = {\rm diag}(10,1)$. Therefore, the
log prior probability we endow on $\theta$ is
\begin{equation}
\log p(\theta) = \log \left(\frac{1}{2\pi\sqrt{10}}\right) - \frac{1}{2}\theta^T\Sigma^{-1}\theta.
\end{equation}
To generate the data, we use the following Gaussian mixture with tied means:
\begin{equation}\label{eq:x_points}
x_i \sim \frac{1}{2}\mathcal{N}(\theta_1, \sigma_x^2) + \frac{1}{2}\mathcal{N}(\theta_1+\theta_2, \sigma_x^2)
\end{equation}
where, again following~\cite{langevin_2011}, we set $\sigma_x^2 = 2$. This means
the log likelihood of a single data instance is
\begin{equation}
\log p(x_i \mid \theta) = \log \frac{1}{4\sqrt{\pi}} +
\log\left(\exp\left(-\frac{1}{4}(x_i - \theta_1)^2\right) + \exp\left(-\frac{1}{4}(x_i -
(\theta_1+\theta_2))^2\right)\right)
\end{equation}
The official problem statement is as follows: given some number of conditionally
independent data points $x_1, x_2, \ldots, x_N$ generated according
to~(\ref{eq:x_points}), determine the posterior distribution of $\theta$:
\begin{equation}\label{eq:log_post}
\log p(\theta \mid x_1,\ldots,x_N) = \log p(\theta) + \sum_{i=1}^N\log p(x_i \mid \theta).
\end{equation}
Alternatively, depending on the application, we may opt to instead pick a point
estimate\footnote{We do this for the logistic regression experiment, described
in Section~\ref{ssec:logistic}.} of $\theta$, such as the MAP estimate. (If $N$
is extremely large, it will cause the posterior to peak sharply at its modes,
reducing distribution estimates to point estimates.)

@@

% Daniel: This was some stuff from the Logistic Regression appendix, but we no
% longer need that.

%% In the logistic regression experiment of Section~\ref{ssec:logistic}, the
%% feature vector for an image consists of its pixel values, normalized between 0
%% and 1. For simplicity, we consider the binary classification case, so we only
%% use digits one ($y = 1$) and seven ($y = -1$).  The probability of the $i^{\rm
%% th}$ output $y_i \in \{-1,+1\}$ with the input vector $x_i$ is
%% \begin{equation}\label{eq:logistic_probability}
%% p(y_i \mid x_i) = \frac{1}{1+\exp (-y_i \theta^T x_i)},
%% \end{equation}
%% where $\theta$ is the 784-length parameter vector.
